Fail Faster

### 2025-03-22 paper/paper.tex
wibbling

This is due to two major sources of inefficiency.
First, the same high-level design approach that gives generator libraries their
flexibility also introduces layered abstractions and indirections that are difficult for
compilers to optimize, shifting work to run time.
Second, the \rand that a generator uses to make choices can
constitute an unexpectedly high proportion of its runtime. These factors combine to limit 
the effectiveness of PBT. \jwc{Unsure if ``constitute a high proportion of its runtime'' limits effectiveness. it's more that the choice \rand limits effectiveness.}
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi

We study and address these issues separately.
First, we show how to use \textit{multi-stage programming}, a classic technique for 
building fast DSLs “without regret”~\cn{}, to eliminate the overhead of 
many common generator abstractions. We refer to this approach as \name.
% Secondly, we demonstrate the
% impact of diferent choices of randomness library on generator performance by conducting a
% controlled comparison.
Second, we demonstrate that a DSL's choice of \rand{} has a
dramatic impact on its performance by performing a controlled experiment,
isolating the performance benefits of faster randomness in a clearly quantifiable manner.
\jwc{Another cut at this.}

@@

In summary, we show that PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions. \jwc{needs a bit about RNG --- BCP thoughts?}
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiency in PBT generator libraries, which can significantly impact performance:
    abstraction overhead and \rand choice.
    \item We present \name{}, a technique for eliminates the abstraction overhead of generators. We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
    \item We demonstrate that writing generators using \name{} and fast \rands{} yields performance improvements.
    We precisely quantify the impact of each optimization separately, and explain how they play distinct yet complementary roles in improving generator performance.
    Finally, we show that these performance improvements extend to significantly improved bug-finding speed, finding some bugs as much as \todo{\%\%} faster.
\end{enumerate}

>>>>

This is due to two major sources of inefficiency.
First, the same high-level design approach that gives generator libraries their
flexibility also introduces layered abstractions and indirections that are difficult for
compilers to optimize, shifting work to run time.
Second, the \rand that a generator uses to make choices can
constitute an large proportion of its running time, magnifying the impact of an inefficient one.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi

In this paper, we study and address these issues.
First, we show how to use \textit{multi-stage programming}, a classic technique for 
building fast DSLs “without regret”~\cn{}, to eliminate the overhead of 
many common generator abstractions. We refer to this approach as \name.
% Secondly, we demonstrate the
% impact of diferent choices of randomness library on generator performance by conducting a
% controlled comparison.
Second, we demonstrate that a DSL's choice of \rand{} has a
dramatic impact on its performance by performing a controlled experiment,
isolating the performance benefits of faster randomness in a clearly quantifiable manner.
\jwc{Another cut at this.}

@@

In summary, we show that PBT generator DSLs incur significant performance costs across languages, and present a general technique for improving their efficiency while preserving their idiomatic abstractions. \jwc{needs a bit about RNG --- BCP thoughts?}
We make the following contributions:
\begin{enumerate}
    \item We identify two key sources of inefficiency in PBT generator libraries, which can significantly impact performance:
    abstraction overhead and \rand choice.
    \item We present \name{}, a technique for eliminates the abstraction overhead of generators. We implement \name{} in both OCaml and Scala 3, showcasing its cross-language applicability.
    \item We demonstrate that writing generators using \name{} and fast \rands{} yields significant performance improvements.
    We precisely quantify the impact of each optimization separately, and explain how they play distinct yet complementary roles in improving generator performance.
    Finally, we show that these performance improvements extend to improved bug-finding speed, finding some bugs as much as \todo{\%\%} faster.
\end{enumerate}
